What paradigms can webcam eye-tracking be used for? Attempted replications of five ``classic'' cognitive science experiments

### 2025-03-18 manuscript.Rmd
minor revisions

A robust solution to all these problems is online experiments, particularly with volunteer citizen scientists as participants [@li2024developmental; @hartshorne2019thousand; @reinecke2015labinthewild; @gosling2010wired]. Historically, this option has not been available for eye-tracking, since few potential subjects have expensive eyetrackers at home. In principal, this could be circumvented by using so-called "poor-man's" eye-tracking, in which researchers videotape the subject's face and code eye position frame-by-frame [@snedeker2003using]. Most computers now have high-quality built-in cameras, so recording video is straightforward. Unfortunately, coding images frame-by-frame is extremely time-intensive, making large-sample studies unrealistic. However, in recent years, image analysis has improved to the point where this work can be automated with reasonable accuracy [@papoutsaki2016webgazer; @zheng2018rapid; @burton2014comparison; @skovsgaard2011evaluation]. Nonetheless, webcam-based eyetracking only started to be used regularly in research with the advent of `Webgazer.js` [@papoutsaki2016webgazer], a webcam-based Javascript plug-in that works in the browser and which can be integrated with any Javascript web interface, including `jsPsych` [@deleeuwJsPsychJavaScriptLibrary2015], `Gorilla` [@anwyl2020gorilla], or `lab.js` [@henninger2021lab].

@@

While promising, there are some salient limitations. First, many of the studies report effects that are smaller or later than what had been previously observed in the lab [@slim2022moving; @degen2021seeing; @van2023validation]. Importantly, accurate timing on a web browser is not trivial [@passell2021cognitive; @de2016psychophysics], and subtle programming choices can significantly affect the accuracy of `WebGazer.js` timing [@yang2021webcam].^[See also discussion at https://github.com/jspsych/jsPsych/discussions/1892] Since most of the prior work did not address these timing issues, it is not clear how many of the reported lab/web differences would resolve.

Second, prior work has focused on studies with relatively coarse-grained regions of interest [but see @semmelmann2018online], dividing the screen either in half or in quadrants. This is particularly salient with [@prystauka2023online], which simplified @altmannIncrementalInterpretationVerbs1999's design so that regions of interest are quadrants rather than the finer-grained ROIs used in the original. Certainly, webcam eyetracking will not be as spatially fine-grained as an infrared eyetracker, but we do not yet have a good sense of the limits.

@@

Second, researchers need to take care with timing accuracy. Running studies on modern computers requires synchronizing multiple clocks, something that does not happen automatically. Failure to synchronize can result in substantial inaccuracies [@slim2022moving]. As indicated by the current work, the current jsPsych implementation of WebGazer.js has reasonably good timing synchronization. We recommend that future researchers use jsPsych or another package with similar documented precision.

>>>>

A robust solution to all these problems is online experiments, particularly with volunteer citizen scientists as participants [@li2024developmental; @hartshorne2019thousand; @reinecke2015labinthewild; @gosling2010wired]. Historically, this option has not been available for eye-tracking, since few potential subjects have expensive eyetrackers at home. In principal, this could be circumvented by using so-called "poor-man's" eye-tracking, in which researchers videotape the subject's face and code eye position frame-by-frame [@snedeker2003using]. Most computers now have high-quality built-in cameras, so recording video is straightforward. Unfortunately, coding images frame-by-frame is extremely time-intensive, making large-sample studies unrealistic. However, in recent years, image analysis has improved to the point where this work can be automated with reasonable accuracy [@papoutsaki2016webgazer; @zheng2018rapid; @burton2014comparison; @skovsgaard2011evaluation]. Nonetheless, webcam-based eyetracking only started to be used regularly in research with the advent of `Webgazer.js` [@papoutsaki2016webgazer], a webcam-based Javascript plug-in that works in the browser and which can be integrated with any Javascript web interface, including `jsPsych` [@deleeuwJsPsychJavaScriptLibrary2015], `Gorilla` [@anwyl2020gorilla], `lab.js` [@henninger2021lab], or `PsychoJS` [@peirce2019psychopy2].

@@

While promising, there are some salient limitations. First, many of the studies report effects that are smaller or later than what had been previously observed in the lab [@van2024validation; @slim2022moving; @degen2021seeing; @van2023validation; @bogdan2024investigating]. Importantly, accurate timing on a web browser is not trivial [@passell2021cognitive; @de2016psychophysics], and subtle programming choices can significantly affect the accuracy of `WebGazer.js` timing [@yang2021webcam].^[See also discussion at https://github.com/jspsych/jsPsych/discussions/1892] Since most of the prior work did not address these timing issues, it is not clear how many of the reported lab/web differences would resolve.

Second, prior work has focused on studies with relatively coarse-grained regions of interest [but see @semmelmann2018online; @bogdan2024investigating], dividing the screen either in half or in quadrants. This is particularly salient with [@prystauka2023online], which simplified @altmannIncrementalInterpretationVerbs1999's design so that regions of interest are quadrants rather than the finer-grained ROIs used in the original. Certainly, webcam eyetracking will not be as spatially fine-grained as an infrared eyetracker, but we do not yet have a good sense of the limits.

@@

Second, researchers need to take care with timing accuracy. Running studies on modern computers requires synchronizing multiple clocks, something that does not happen automatically. Failure to synchronize can result in substantial inaccuracies [@slim2022moving]. As indicated by the current work, the current jsPsych implementation of WebGazer.js has reasonably good timing synchronization. While it is possible that other Web-based eyetracking software achieves (or can achieve) the same level of precision as currently offered by jsPsych, we urge researchers to be cautious about using software for which the precision has not been documented.
